Aneuploidy can be an evolutionary detour \\ on the path to adaptation

### 2022-12-14 ms/ms.tex
shorter abstract

% Abstract 250 words
\begin{abstract} 
Aneuploidy is common in eukaryotes, often leading to decreased growth and fitness. However, evidence from yeast and other fungi, as well as human tumour cells, suggests that specific aneuploidies can be beneficial under stressful conditions and facilitate  adaptation.
In a prominent example, an evolutionary experiment with yeast, populations evolving under heat stress had become aneuploid (chromosome III), only to later revert back to euploid after genetic mutations have accumulated. It has therefore been suggested that aneuploidy serves as a "stepping stone" on the path to adaptation.

Here, we test this hypothesis. First, we apply DNA sequencing to show that mutant alleles common in aneuploid cells are uncommon in the evolved euploid population. 
Second, we develop an evolutionary model with both aneuploidy and  mutation, and fit it to the results of the experiment using a Bayesian inference framework.
We then predict the genotype frequency dynamics during the experiment, demonstrating that the majority of the evolved euploid population likely did not descend from aneuploid cells, but rather directly from the euploid wild-type population.
Our model shows how the beneficial mutation supply--the product of population size and beneficial mutation rate--determine the evolutionary dynamics: with a low mutation supply, a large fraction of the evolved population may descend from aneuploid cells; but with a high mutation supply, beneficial mutations are generated before fixation of aneuploidy, and can outcompete aneuploidy due to the latter's inherent fitness cost.

Together, our results suggest that despite its potential fitness benefits under stress, aneuploidy can be an evolutionary "detour" rather than a "stepping stone": it can delay, rather than facilitate, the adaptation of the population, and cells that become aneuploid may leave less descendants compared to cells that remain diploid.
\end{abstract}

% TODO Significance statement 120 words

>>>>

% Abstract 250 words
\begin{abstract} 
Aneuploidy is common in eukaryotes, often leading to decreased growth and fitness. However, evidence from yeast and other fungi, as well as human tumour cells, suggest that specific aneuploidies can be beneficial under stressful conditions and facilitate adaptation.
In a prominent experimental evolutionary study with yeast, populations evolving under heat stress became aneuploid, only to later revert back to euploidy after genetic mutations have accumulated. It has therefore been suggested that aneuploidy serves as a "stepping stone" on the path to adaptation.
Here, we test this hypothesis. First, we apply DNA sequencing to show that mutant alleles common in aneuploid cells are uncommon in the evolved euploid population. 
Second, we develop an evolutionary model with both aneuploidy and mutation and fit it to the experimental results using Bayesian inference. We then predict the genotype frequency dynamics during the experiment, demonstrating that the majority of the evolved euploid population likely did not descend from aneuploid cells, but rather from the euploid wildtype population.
Together, our results suggest that despite its potential fitness benefits under stress, aneuploidy can be an evolutionary "detour" rather than a "stepping stone": it can delay, rather than facilitate, the adaptation of the population, and cells that become aneuploid may leave less descendants compared to cells that remain diploid.
\end{abstract}
